Cubical Types with Positive and Negative Polarities

### 2014-05-19 ring.tex
separate syntax from denotation

\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

@@

We use $\tau$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\cubt$ for
general $n$-dimensional cubical types. An $n$-dimensional cubical type
$\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and a
negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical types of a lower dimension; the 1-dimensional case reduces to plain
first-order types. As we explain below, cubical types are closed under sums,
differences, products, and can express higher-order functions $\cubt_1 \lolli
\cubt_2$ as $(\ominus \cubt_1) \oplus \cubt_2$. The syntax of the types is
summarized below:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \\
\\
\cubt &::=& \tau \alt \nodet{\cubt_1}{\cubt_2} 
      \alt \cubt \oplus \cubt 
      \alt \cubt \otimes \cubt 
      \alt \ominus \cubt 
\end{array}\]
Examples of 1-dimensional cubical types include all the regular first-order
types $\tau$. A 2-dimensional cubical type looks like
$\nodet{\tau_1}{\tau_2}$ which intuitively corresponds to the difference
$\tau_1 - \tau_2$ of the two types. A 3-dimensional cubical type looks like
$\nodet{(\nodet{\tau_1}{\tau_2})}{(\nodet{\tau_3}{\tau_4})}$ which
intuitively corresponds to the iterated difference of types
$(\tau_1-\tau_2)-(\tau_3-\tau_4)$ where the successive colors from the root
encode the sign as follows: white in the outermost box corresponds to a
positive polarity and gray in the outermost box corresponds to a negative
polarity; then, as we move inwards the polarities alternate for every change
of color. It is possible for cubical types not be balanced in general but it
is always possible to extend the shorter subtree to an arbitrary depth using
the fact (to be justified later) that~$\tau$ is equivalent to
$\nodet{\tau}{0}$.

We now explain how cubical types are closed under sum, difference, and
products. In the recursive definitions below, we assume that the types have
been extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~\tau &=& \tau \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
\tau_1 \oplus \tau_2 &=& \tau_1 + \tau_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
\tau_1 \otimes \tau_2 &=& \tau_1 * \tau_2 \\
\tau \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{\tau \otimes \cubt_1}{\tau \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

\begin{figure*}
\[\begin{array}{c}
\nodet{\tau_1}{\tau_2}
\quad\otimes\quad
\nodet{(\nodet{\tau_3}{\tau_4})}{(\nodet{\tau_5}{\tau_6})} = 
\\
\nodet{(\nodet{{(\nodet{\tau_1*\tau_3}{\tau_1*\tau_4})}}
              {{(\nodet{\tau_1*\tau_5}{\tau_1*\tau_6})}})}
      {(\nodet{{(\nodet{\tau_2*\tau_3}{\tau_2*\tau_4})}}
              {{(\nodet{\tau_2*\tau_5}{\tau_2*\tau_6})}})}
\end{array}\]
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

A few examples might provide some needed intuition. First we note that the
general definitions of sum and product reduce to the conventional sums and
products for 1-dimensional types. Also the sum of 2-dimensional types reduces
to the sum used in the \textbf{Int} construction. The interesting new
development is the product definition. The example in Fig.~\ref{mult} shows
that in general the product of an $n$-dimensional type with an
$m$-dimensional type gives an $(n+m)$-dimensional type. For low dimensions
such as the ones used in the example above, it is possible to visualize what
is happening in coordinate space (see Fig.~\ref{multv}).

\begin{figure*}
\begin{tikzpicture}
\node[left] at (0,0) {$\tau_1^+$};
\node[right] at (0.6,0) {$\tau_2^-$};
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$}; 
%%
\node[below left] at (2.5,-0.5) {$\tau_3^+$};
\node[below right] at (3.5,-0.5) {$\tau_4^-$};
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[above left] at (2.5,0.5) {$\tau_6^-$};
\node[above right] at (3.5,0.5) {$\tau_5^+$};
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (4.5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(\tau_2*\tau_3)^-$};
\node[right] at (9.5,0.75) {$(\tau_2*\tau_4)^+$};
\node[above right] at (10.2,1.2) {$(\tau_2*\tau_5)^-$};
\node[above left] at (8.2,1.2) {$(\tau_2*\tau_6)^+$};
%%
\node[left] at (7.5,-0.75) {$(\tau_1*\tau_3)^+$};
\node[right] at (9.5,-0.75) {$(\tau_1*\tau_4)^-$};
\node[above right] at (10.2,-0.3) {$(\tau_1*\tau_5)^+$};
\node[above] at (8.2,-0.3) {$(\tau_1*\tau_6)^-$};
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical types.}
\end{figure*}

>>>>

\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\denc}[1]{\llbracket #1 \rrbracket^c}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

@@

We use $t$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types.  We use $\tau$ for cubical
types: these have the same formal structure as regular types but also include
\emph{negative} types. The syntax of the types is summarized below:
\[\begin{array}{rcl}
t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
\tau &::=& t \alt 
      \alt \tau_1 + \tau_2
      \alt \tau_1 * \tau_2
      \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.

It is easy to find a model for regular types: finite sets are such a
model. In more detail:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \bullet \} \\
\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of types maps to disjoint union of sets. The product of types maps to the
cartesian product of sets.

For cubical types, their denotations are $n$-dimensional cubical sets with
polarities, written as $\nodet{\cubt_1}{\cubt_2}$ An $n$-dimensional cubical
set $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubical set of a lower dimension; the 1-dimensional case reduces to plain
sets (which are the denotations of first-order types). As we explain below,
cubical sets are closed under sums, differences, products.

Examples of 1-dimensional cubical sets include all the finite sets $S$
denoting regular first-order types $t$. A 2-dimensional cubical set looks
like $\nodet{S_1}{S_2}$ which intuitively corresponds to the difference $t_1
- t_2$ of the two types whose denotations are $S_1$ and $S_2$ respectively. A
3-dimensional cubical type looks like
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the successive colors from the root encode the
sign as follows: white in the outermost box corresponds to a positive
polarity and gray in the outermost box corresponds to a negative polarity;
then, as we move inwards the polarities alternate for every change of
color.

We now give the denotation of cubical types:
\[\begin{array}{rcl}
\denc{t} &=& \den{t} = \nodet{\den{t}}{0} = 
  \nodet{\nodet{\den{t}}{0}}{0} = \ldots \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible for cubical types not be balanced in general but it is always
possible to extend the shorter subtree to an arbitrary depth using the first
clause which formalizes that $t$ is equivalent to $t-0$ and to $(t-0)-0$ etc.
In the recursive definitions below, we assume that the types have been
extended to be balanced as needed:
\[\begin{array}{rcl}
\ominus~\tau &=& \tau \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
\tau_1 \oplus \tau_2 &=& \tau_1 + \tau_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
\tau_1 \otimes \tau_2 &=& \tau_1 * \tau_2 \\
\tau \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{\tau \otimes \cubt_1}{\tau \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

\begin{figure*}
\[\begin{array}{c}
\nodet{S_1}{S_2}
\quad\otimes\quad
\nodet{(\nodet{S_3}{S_4})}{(\nodet{S_5}{S_6})} = 
\\
\nodet{(\nodet{{(\nodet{S_1 \times S_3}{S_1 \times S_4})}}
              {{(\nodet{S_1 \times S_5}{S_1 \times S_6})}})}
      {(\nodet{{(\nodet{S_2 \times S_3}{S_2 \times S_4})}}
              {{(\nodet{S_2 \times S_5}{S_2 \times S_6})}})}
\end{array}\]
\caption{\label{mult}Example of multiplication of two cubical types.}
\end{figure*}

The definition of negation is natural: it recursively swaps the positive and
negative subspaces. We note that the general definitions of sum and product
reduce to the conventional sums and products for 1-dimensional types. Also
the sum of 2-dimensional types reduces to the sum used in the \textbf{Int}
construction. The interesting new development is the product definition. The
example in Fig.~\ref{mult} shows that in general the product of an
$n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

\begin{figure*}
\begin{tikzpicture}
\node[left] at (0,0) {$S_1^+$};
\node[right] at (0.6,0) {$S_2^-$};
\draw[-] (0,0) -- (0.6,0);
\node at (1.6,0) {$\otimes$}; 
%%
\node[below left] at (2.5,-0.5) {$S_3^+$};
\node[below right] at (3.5,-0.5) {$S_4^-$};
\draw[-] (2.5,-0.5) -- (3.5,-0.5);
\draw[-] (2.5,-0.5) -- (2.5,0.5);
\node[above left] at (2.5,0.5) {$S_6^-$};
\node[above right] at (3.5,0.5) {$S_5^+$};
\draw[-] (2.5,0.5) -- (3.5,0.5);
\draw[-] (3.5,-0.5) -- (3.5,0.5);
%% 
\node at (4.5,0) {$=$};
%% 
\node[left] at (7.5,0.75) {$(S_2 \times S_3)^-$};
\node[right] at (9.5,0.75) {$(S_2 \times S_4)^+$};
\node[above right] at (10.2,1.2) {$(S_2 \times S_5)^-$};
\node[above left] at (8.2,1.2) {$(S_2 \times S_6)^+$};
%%
\node[left] at (7.5,-0.75) {$(S_1 \times S_3)^+$};
\node[right] at (9.5,-0.75) {$(S_1 \times S_4)^-$};
\node[above right] at (10.2,-0.3) {$(S_1 \times S_5)^+$};
\node[above] at (8.2,-0.3) {$(S_1 \times S_6)^-$};
%%
\draw[-] (7.5,0.75) -- (9.5,0.75);
\draw[-] (9.5,0.75) -- (10.2,1.2);
\draw[-] (10.2,1.2) -- (8.2,1.2);
\draw[-] (8.2,1.2) -- (7.5,0.75);
%%
\draw[-] (7.5,-0.75) -- (9.5,-0.75);
\draw[-] (9.5,-0.75) -- (10.2,-0.3);
\draw[-,dashed] (10.2,-0.3) -- (8.2,-0.3);
\draw[-,dashed] (8.2,-0.3) -- (7.5,-0.75);
%%
\draw[-] (7.5,0.75) -- (7.5,-0.75);
\draw[-] (9.5,0.75) -- (9.5,-0.75);
\draw[-] (10.2,1.2) -- (10.2,-0.3);
\draw[-,dashed] (8.2,1.2) -- (8.2,-0.3);
\end{tikzpicture}
\caption{\label{multv}Visualization of multiplication of two cubical types.}
\end{figure*}
